---
title: "R Notebook"
output:
 html_notebook:
  code_folding: hide
---




## Exercise 1

Navigate to the directory with tutorials. Open exercises.nb.html in a browser. You are in the middle of the first exercise:).
Start RStudio session and navigate to the tutorials directory. 


## Exercise 2

Open Jupyter notebook `intro_to_python.ipynb`. 

Choose from exercises 2-4 below according to your needs. We will follow together exercises chosen according to a vote, but trainers will be going around and help also if you have chosen a different exercise.
You will find a cheatsheet for jupyter notebook in the cheatsheets. 

```{bash }
# commandline: navigate to your working directory. Adjust the command below to reflect the path to .ipynb file
jupyter notebook intro_to_python.ipynb 

```


At the bottom of the notebook, add a new section 'Exercise during the course'. 
Load the package pandas and read in the `20-07-25patients_metadatav2.csv` file with patient metadata. 

```{python}
import pandas as pd

file = 'R_intro/working_dir/data/20-07-25patients_metadatav2.csv'

patient_data = pd.read_csv(file)
```

Explore its content and available methods. What are the column names? Have a peek on the top few rows. 
```{python}
#have a look on top/bottom rows 
patient_data.head()
patient_data.tail()

#List column names
patient_data.columns

#Access subset of columns by row/column number or name
patient_data.iloc[1:3,[4,2,6]]
patient_data.loc[:,['patient_id','sex','age']]
```

Take a subset including only females with low disease severity
```{python}

patient_data.loc[(patient_data.class_dss = = 'Low') & (patient_data.sex = = 'F')]

```

Plot a histogram of WHO scores at admission (`covid_ip_bleed_01_WHO_score`)
```{python}
patient_data.hist('covid_ip_bleed_01_WHO_score')
```


Plot a boxplot of admission WHO scores by sex
```{python}
patient_data.boxplot(column = 'peak_WHO_ordinal_scale',by = 'sex',color = 'red')

```


## Exercise 3
Use R and tidyverse to explore covid data. 
Read in the metadata (`20-07-25patients_metadatav2.csv`) and tests results (online at `https://www.immunophenotype.org/wp-content/uploads/2020/07/2020-06-30flow_sero_cyto__ifnaexport.csv`).  

```{r}
metadata_loc = "data/20-07-25patients_metadatav2.csv"
patient_info <- read_csv(metadata_loc)

tests_loc = "https://www.immunophenotype.org/wp-content/uploads/2020/07/2020-06-30flow_sero_cyto__ifnaexport.csv" 
covid_tests <- read_csv(tests_loc) 

## Just in case you cannot access it online, the downloaded data is in tests_loc = 'R_intro/data/just_in_case/covid_just_in_case.csv.gz' (change tests_loc to this value)
```



Explore the metadata and tests objects. How many patients were in the study? How many samples per patient? For how many patients BMI info is available? 
Below are coding examples of how to obtain this information, but there are many ways of doing it.
```{r }

dim(covid_tests)
covid_tests

#How many patients were in the study?
patient_info
patient_info$patient_id %>%
 unique%>%
 length  

#How many samples per patient? 
covid_tests %>%
 group_by(patient_id)%>%
 summarise(samples_per_patient = n())

#For how many patients BMI info is available?  
patient_info%>%
 summarise(sum(!is.na(BMI)))
 


```

Now add the test data. 
Does the number of samples per patient depend on the severity of the disease?

```{r}
covid_data_annotated <- covid_tests %>%
 left_join( patient_info, ., by = c( "patient_id" ) )

covid_data_annotated %>%
 group_by(patient_id, class_dss)%>%
 summarise(samples_per_patient = n()) %>%
 ggplot(aes(x = class_dss, y = samples_per_patient, col = class_dss))+
  geom_boxplot() 


library(ggbeeswarm)
 covid_data_annotated %>%
 group_by(patient_id, class_dss)%>%
 summarise(samples_per_patient = n()) %>%
 ggplot(aes(x = factor(class_dss, levels = c("LRTI_nonCovid", "seroneg_control","seropos_control" ,"Low", "Moderate","Severe")),
       y = samples_per_patient, col = class_dss))+
  geom_boxplot() +
  geom_quasirandom(varwidth = TRUE,grouponX = TRUE, alpha = 0.5, size = 3)+
  theme_classic()+
  ylim(0,8)+ ylab("Samples per patient")+
  xlab("Disease level")+
  guides(colour = "none")+
  coord_flip()
 
```

For patients with more than 1 sample, plot how their RBD_IgG_Norm_sero and lymphocyte count change in time.
```{r}

patients_with_more_than_1_sample <- covid_data_annotated %>%
 group_by(patient_id, class_dss)%>%
 summarise(samples_per_patient = n()) %>%
 filter(samples_per_patient > 1)%>%
 .$patient_id


covid_data_annotated %>%
 filter(patient_id %in% patients_with_more_than_1_sample)%>%
 separate(Clinical_sample,into = c("patient","consecutive_sample"),sep = "n")%>%
 ggplot(aes(x = consecutive_sample, y = RBD_IgG_Norm_sero, group = patient_id, col = class_dss))+
 geom_point()+ geom_line() +
 scale_colour_discrete(name = "severity group")+
 theme_classic() 

another_column = 'Time_06/Cells_06/Singlets1_06/Singlets2_06/CD45p_06/Lymphocytes_06 | Count_back'

covid_data_annotated %>%
 filter(patient_id %in% patients_with_more_than_1_sample)%>%
 separate(Clinical_sample,into = c("patient","consecutive_sample"),sep = "n")%>%
 ggplot(aes(x = consecutive_sample, y = get(another_column), group = patient_id, col = class_dss))+
 geom_point()+ geom_line() +
 scale_colour_discrete(name = "severity group")+
 theme_classic() 

```



## Exercise 4

The goal of this exercise is to have a go at using a package you have not used before. If you know ggbio - maybe choose another package form Bioconductor, install it and have a go on accessing package-specific objects, related functions and methods.  
Load the plotting library (ggbio) and the library with human annotation:
```{r}
library(ggbio) 
library(Homo.sapiens) 
```

Plot IL7 transcripts:
```{r}
data(genesymbol, package = "biovizBase") #what sort of object is genesymbol?
gr_il7 <- genesymbol[c("IL7")]    

#Try your own gene. What sort of object is gr_il7?
#What does it mean 'seqinfo: 45 sequences from an unspecified genome; no seqlengths'? Have a look on slots in gr_il7
just_region <- range(gr_il7, ignore.strand = TRUE)
p.txdb <- autoplot(Homo.sapiens, which = just_region) 
p.txdb+theme_bw()
autoplot(Homo.sapiens, which = gr_il7) 
```


Make a similar plot for genes located in chr19:39327059-39342999 or chr4:123279159-123414842
```{r}
gr <- GRanges(
  seqnames = Rle(c("chr4"),1),
  ranges = IRanges(123279159, end = 123414842, names = "some_region"),
  strand = Rle(strand(c("*")), c(1)))

 autoplot(Homo.sapiens, which = gr, label.color = "red")
```

Write a function which takes in a data frame with chromosome location and plot transcripts across this area.


What is going under the hood of autoplot? How it knows how to plot Homo.sapiens content?

```{r}
showMethods(autoplot)
class(Homo.sapiens)

getMethod(autoplot,signature = "OrganismDb")
```



